Air Quality Data from EPA AQS

### 2021-02-17 code/data-analysis/factor-analysis.Rmd
smoothing spline for apportioned mass - new mass app combination

```{r Load datasets}
load(here('data','processed', 'pmc-daily.RDa'))
load(here('data','processed', 'aqs-sites.RDa'))
```

@@

### high load
high.load <- lmer(data = dt.pmc.daily,
              PM ~ SI + PB + OC + K + SO4 + CL + NI + FE + (1|Site.ID))
tab_model(high.load, show.se = T, show.ci = F, show.aic = T,
            show.re.var = F, show.icc = T, show.obs = T, show.ngroups = T,
          title = "Higher Load Tracers")

### high health impact
high.health <- lmer(data = dt.pmc.daily,
              PM ~ SI + PB + NO3 + K + SO4 + CL + NI + FE + (1|Site.ID))
tab_model(high.health, show.se = T, show.ci = F, show.aic = T, digits = 2,
            show.re.var = F, show.icc = T, show.obs = T, show.ngroups = T,
          title = "Higher Health Impact Tracers")
# Nagative Ni

### high health impact ii
high.health <- lmer(data = dt.pmc.daily,
              PM ~ SI + PB + NO3 + K + SO4 + CL + NI + MN + (1|Site.ID))
tab_model(high.health, show.se = T, show.ci = F, show.aic = T, digits = 2,
            show.re.var = F, show.icc = T, show.obs = T, show.ngroups = T,
          title = "Higher Health Impact Tracers - II")

@@

```{r - Single Tracer Mass Apportionment}
dt.sites <- dt.aqs.sites %>% 
  transmute(Site.ID,
            Latitude,
            Longitude,
            Location = ifelse(Location.Setting != "RURAL", "Urban", "Rural"),
            Region = ifelse(Region.IV == 1, "West",
                            ifelse(Region.IV == 2, "Midwest",
                                   ifelse(Region.IV == 3, "South", "Northeast"))))

@@

so.coef <- function(model, cmp){
  source = as.numeric(summary(model)$coefficients[,1][cmp])
  return(source)
}

dt.mass.app <- dt.pmc.daily %>% 
  mutate(Soil = so.coef(high.load, 'SI') * SI,
         Metal = so.coef(high.load, 'PB') * PB,
         Traffic = so.coef(high.load, 'OC') * OC,
         Biomass = so.coef(high.load, 'K') * K,
         Coal = so.coef(high.load, 'SO4') * SO4,
         Salt = so.coef(high.load, 'CL') * CL,
         Oil = so.coef(high.load, 'NI') * NI,
         Steel = so.coef(high.load, 'FE') * FE) %>% 
  select(Site.ID, Date, PM, Sources) %>% 
  inner_join(dt.sites) %>% 
  pivot_longer(cols = c('PM', Sources),
               names_to = "Source",
               values_to = "Mass")

save(dt.mass.app, file = here('data', 'processed', 'mass-apportioned.RDa'))

mass.summary.us <- dt.mass.app %>% 
  group_by(Source) %>% 
  summarise(Mass.mean = round(mean(Mass),2),
            Mass.sd = round(sd(Mass),2)) %>% 
  mutate(Unit = "US")

mass.summary.region <- dt.mass.app %>% 
  group_by(Source, Region) %>% 
  summarise(Mass.mean = round(mean(Mass),2),
            Mass.sd = round(sd(Mass),2)) %>% 
  rename(Unit = Region)

mass.summary.locset <- dt.mass.app %>% 
  group_by(Source, Location) %>% 
  summarise(Mass.mean = round(mean(Mass),2),
            Mass.sd = round(sd(Mass),2)) %>% 
rename(Unit = Location)

mass.summary <- mass.summary.us %>% 
  add_row(mass.summary.region) %>% 
  add_row(mass.summary.locset)

write.csv(mass.summary,
          file = here('output','table', 'mass-app-summary.csv'))

dt.mass.map <- dt.mass.app %>% 
  group_by(Site.ID, Source) %>% 
  summarise(Mass = mean(Mass)) %>% 
  pivot_wider(names_from = "Source",
              values_from = "Mass") %>% 
  inner_join(dt.sites)

save(dt.mass.map, file = here('data', 'processed', 'mass-map.RDa'))
write.csv(dt.mass.map, file = here('output','table', 'mass-map.csv'), row.names = F)
```

>>>>

```{r Load datasets}
load(here('data','processed', 'pmc-daily.RDa'))
load(here('data','processed', 'aqs-sites.RDa'))
load(here('data','processed', 'mass-apportioned.RDa'))
```

@@

### high load -- Comb1
high.load <- lmer(data = dt.pmc.daily,
              PM ~ SI + PB + OC + K + SO4 + CL + NI + FE + (1|Site.ID))
tab_model(high.load, show.se = T, show.ci = F, show.aic = T,
            show.re.var = F, show.icc = T, show.obs = T, show.ngroups = T,
          title = "Higher Load Tracers")

### high health impact ii - Comb2
high.health <- lmer(data = dt.pmc.daily,
              PM ~ SI + PB + NO3 + K + SO4 + CL + NI + MN + (1|Site.ID))
tab_model(high.health, show.se = T, show.ci = F, show.aic = T, digits = 2,
            show.re.var = F, show.icc = T, show.obs = T, show.ngroups = T,
          title = "Higher Health Impact Tracers - II")
# CL non sig

@@

```{r - Single Tracer Mass Apportionment - function}
dt.sites <- dt.aqs.sites %>% 
  transmute(Site.ID,
            Latitude,
            Longitude,
            Location = ifelse(Location.Setting != "RURAL", "Urban", "Rural"),
            Region = ifelse(Region.IV == 1, "West",
                            ifelse(Region.IV == 2, "Midwest",
                                   ifelse(Region.IV == 3, "South", "Northeast"))))

@@

so.coef <- function(model, cmp){
  source = as.numeric(summary(model)$coefficients[,1][cmp])
  return(source)
}
```

```{r - Single Tracer Mass Apportionment - Combination 1}
dt.mass.app.c1 <- dt.pmc.daily %>% 
  mutate(Soil = so.coef(high.load, 'SI') * SI,
         Metal = so.coef(high.load, 'PB') * PB,
         Traffic = so.coef(high.load, 'OC') * OC,
         Biomass = so.coef(high.load, 'K') * K,
         Coal = so.coef(high.load, 'SO4') * SO4,
         Salt = so.coef(high.load, 'CL') * CL,
         Oil = so.coef(high.load, 'NI') * NI,
         Steel = so.coef(high.load, 'FE') * FE) %>% 
  select(Site.ID, Date, PM, Sources) %>% 
  inner_join(dt.sites) %>% 
  pivot_longer(cols = c('PM', Sources),
               names_to = "Source",
               values_to = "Mass")

save(dt.mass.app.c1, file = here('data', 'processed', 'mass-apportioned-c1.RDa'))

mass.summary.us <- dt.mass.app.c1 %>% 
  group_by(Source) %>% 
  summarise(Mass.mean = round(mean(Mass),2),
            Mass.sd = round(sd(Mass),2)) %>% 
  mutate(Unit = "US")

mass.summary.region <- dt.mass.app.c1 %>% 
  group_by(Source, Region) %>% 
  summarise(Mass.mean = round(mean(Mass),2),
            Mass.sd = round(sd(Mass),2)) %>% 
  rename(Unit = Region)

mass.summary.locset <- dt.mass.app.c1 %>% 
  group_by(Source, Location) %>% 
  summarise(Mass.mean = round(mean(Mass),2),
            Mass.sd = round(sd(Mass),2)) %>% 
rename(Unit = Location)

mass.summary.c1 <- mass.summary.us %>% 
  add_row(mass.summary.region) %>% 
  add_row(mass.summary.locset)

write.csv(mass.summary.c1,
          file = here('output','table', 'mass-app-summary-c1.csv'))

dt.mass.map.c1 <- dt.mass.app.c1 %>% 
  group_by(Site.ID, Source) %>% 
  summarise(Mass = mean(Mass)) %>% 
  pivot_wider(names_from = "Source",
              values_from = "Mass") %>% 
  inner_join(dt.sites)

save(dt.mass.map.c1, file = here('data', 'processed', 'mass-map-c1.RDa'))
write.csv(dt.mass.map.c1, file = here('output','table', 'mass-map-c1.csv'), row.names = F)
```

```{r Smoothing Data & Caluculating 1-yr mvavg for each month - Combination 1}
dt.air <- dt.mass.app.c1 %>% 
  transmute(Site.ID,
            Parameter.Symbol = Source,
            Date,
            Value = Mass) %>% 
  arrange(Site.ID, Parameter.Symbol, Date)

dt.mass.app.1yr.c1 <- para.monthly %>% 
  mutate(Year = floor(Date -0.01),
         Month = round((Date - Year)*12)) %>% 
  select(-Date) %>% 
  pivot_wider(names_sort = T,
              names_from = c(Parameter.Symbol),
              values_from = c(para.1yr)) %>% 
  arrange(Site.ID, Year, Month)

save(dt.mass.app.1yr.c1, file = here('data', 'processed', 'mass-app-1yr-c1.RDa'))
write.csv(dt.mass.app.1yr.c1, file = here('data', 'processed', 'mass-app-1yr-c1.csv'), row.names = F)
```

```{r - Single Tracer Mass Apportionment - Combination 2}
dt.mass.app.c2 <- dt.pmc.daily %>% 
  mutate(Soil = so.coef(high.health, 'SI') * SI,
         Metal = so.coef(high.health, 'PB') * PB,
         Traffic = so.coef(high.health, 'NO3') * NO3,
         Biomass = so.coef(high.health, 'K') * K,
         Coal = so.coef(high.health, 'SO4') * SO4,
         Salt = so.coef(high.health, 'CL') * CL,
         Oil = so.coef(high.health, 'NI') * NI,
         Steel = so.coef(high.health, 'MN') * MN) %>% 
  select(Site.ID, Date, PM, Sources) %>% 
  inner_join(dt.sites) %>% 
  pivot_longer(cols = c('PM', Sources),
               names_to = "Source",
               values_to = "Mass")

save(dt.mass.app.c2, file = here('data', 'processed', 'mass-apportioned-c2.RDa'))

mass.summary.us <- dt.mass.app.c2 %>% 
  group_by(Source) %>% 
  summarise(Mass.mean = round(mean(Mass),2),
            Mass.sd = round(sd(Mass),2)) %>% 
  mutate(Unit = "US")

mass.summary.region <- dt.mass.app.c2 %>% 
  group_by(Source, Region) %>% 
  summarise(Mass.mean = round(mean(Mass),2),
            Mass.sd = round(sd(Mass),2)) %>% 
  rename(Unit = Region)

mass.summary.locset <- dt.mass.app.c2 %>% 
  group_by(Source, Location) %>% 
  summarise(Mass.mean = round(mean(Mass),2),
            Mass.sd = round(sd(Mass),2)) %>% 
rename(Unit = Location)

mass.summary.c2 <- mass.summary.us %>% 
  add_row(mass.summary.region) %>% 
  add_row(mass.summary.locset)

write.csv(mass.summary.c2,
          file = here('output','table', 'mass-app-summary-c2.csv'))

dt.mass.map.c2 <- dt.mass.app.c2 %>% 
  group_by(Site.ID, Source) %>% 
  summarise(Mass = mean(Mass)) %>% 
  pivot_wider(names_from = "Source",
              values_from = "Mass") %>% 
  inner_join(dt.sites)

save(dt.mass.map.c2, file = here('data', 'processed', 'mass-map-c2.RDa'))
write.csv(dt.mass.map.c2, file = here('output','table', 'mass-map-c2.csv'), row.names = F)
```

```{r Smoothing Data & Caluculating 1-yr mvavg for each month - Combination 2}
dt.air <- dt.mass.app.c2 %>% 
  transmute(Site.ID,
            Parameter.Symbol = Source,
            Date,
            Value = Mass) %>% 
  arrange(Site.ID, Parameter.Symbol, Date)

dt.mass.app.1yr.c2 <- para.monthly %>% 
  mutate(Year = floor(Date -0.01),
         Month = round((Date - Year)*12)) %>% 
  select(-Date) %>% 
  pivot_wider(names_sort = T,
              names_from = c(Parameter.Symbol),
              values_from = c(para.1yr)) %>% 
  arrange(Site.ID, Year, Month)

save(dt.mass.app.1yr.c2, file = here('data', 'processed', 'mass-app-1yr-c2.RDa'))
write.csv(dt.mass.app.1yr.c2, file = here('data', 'processed', 'mass-app-1yr-c2.csv'), row.names = F)
```
